---
title: "FinalProject_DataCleaning_v1"
author: "Alexandre"
date: "April/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen = 0, digits = 3)
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(bestglm, glmnet, leaps, car, tidyverse, pROC, caret) # add the packages needed
library(xtable)
library(dplyr)
library(readxl)
```


```{r, echo=FALSE}
#clear variables
rm(list=ls())
```

# Geo data

## City names and codes

Source: IBGE, https://www.ibge.gov.br/explica/codigos-dos-municipios.php

```{r}
names <- read.csv("data/city_names_v1.csv", encoding = "UTF-8")
str(names) 

names <- names %>%
  rename(city_code = X.U.FEFF.city_code) %>%
  select(city_code, PT_city_UF)
str(names)   
```


## Codes and names variations

Variations of names, compiled from different data sources used, with corresponding IBGE code

```{r}
names_aux <- read.csv("data/COMPLETE_city_names_v1.csv", encoding = "UTF-8")
str(names_aux) 

names_aux <- names_aux %>%
   rename(PT_city_UF_aux = X.U.FEFF.PT_city_UF_aux) %>%
   select(city_code, PT_city_UF_aux)
str(names_aux)   
```

```{r}
names_aux2 <- read.csv("data/COMPLETE_city_names_v2.csv", encoding = "UTF-8")
str(names_aux2) 

names_aux2 <- names_aux2 %>%
  distinct() %>%
  rename(PT_city_UF_aux = X.U.FEFF.PT_city_UF_aux) %>%
  select(city_code, PT_city_UF_aux)
str(names_aux2)   
```

```{r}
names_aux3 <- read.csv("data/COMPLETE_city_names_v3.csv", encoding = "UTF-8")
str(names_aux3) 

names_aux3 <- names_aux3 %>%
  distinct() %>%
  rename(PT_city_UF_aux = X.U.FEFF.PT_city_UF_aux) %>%
  select(city_code, PT_city_UF_aux)
str(names_aux3)   
```

```{r}
names_aux4 <- read.csv("data/COMPLETE_city_names_v4.csv", encoding = "UTF-8")
str(names_aux4) 

names_aux4 <- names_aux4 %>%
  distinct() %>%
  rename(PT_city_UF_aux = X.U.FEFF.PT_city_UF_aux) %>%
  select(city_code, PT_city_UF_aux)
str(names_aux4)   
```

# Demographic data

## Absolute population (2011-20)

Source: Estimated population, IBGE, table 6576, https://sidra.ibge.gov.br/tabela/6579

```{r}
# Population
pop <- read_xlsx("data/IBGE/tabela6579_pop_2011-20.xlsx", sheet = 1) #local path of database
str(pop)
a <- dim(pop)

pop <- pop[-c(1,2,3,4,a[1]),]

pop <- pop %>%
  rename(city_UF = "Tabela 6579 - População residente estimada",
         pop.2011 = ...2,
         pop.2012 = ...3,
         pop.2013 = ...4,
         pop.2014 = ...5,
         pop.2015 = ...6,
         pop.2016 = ...7,
         pop.2017 = ...8,
         pop.2018 = ...9,
         pop.2019 = ...10,
         pop.2020 = ...11)
#str(pop)

#sum(is.na(pop))

pop <- bind_cols(pop[,1], lapply(pop[,c(2:11)], as.numeric))
str(pop)
```

```{r}
# Calculate annual populational growth
pop <- pop %>%
  mutate(pop.growth2011_20 = (pop.2020/pop.2011)^(1/9)-1)
str(pop)

# Save cleaned data
pop %>%
  write.csv("data_cleaned/db_pop.csv",row.names=FALSE)
```

## Population by age bracket

Source: IBGE, table 2093, https://sidra.ibge.gov.br/tabela/2093#:~:text=Tabela%202093%3A%20Popula%C3%A7%C3%A3o%20residente%20por,Amostra%20%2D%20Caracter%C3%ADsticas%20Gerais%20da%20Popula%C3%A7%C3%A3o

```{r}
# Population by age
pop_age <- read_xlsx("data/IBGE/Censo_tabela2093_pop_idade.xlsx", sheet = 1) #local path of database
#str(pop_age)
a <- dim(pop_age)

pop_age <- pop_age[-c(1,2,3,4,5,a[1]),]

pop_age <- pop_age %>%
  rename(city_UF = "Tabela 2093 - População residente por cor ou raça, sexo, situação do domicílio e grupos de idade - Amostra - Características Gerais da População",
         pop.0_4 = ...5,
         pop.5_9 = ...6,
         pop.10_14 = ...7,
         pop.15_19 = ...9,
         pop.20_24 = ...12,
         pop.25_29 = ...13,
         pop.30_39 = ...14,
         pop.40_49 = ...15,
         pop.50_59 = ...16,
         pop.60_69 = ...17,
         pop.70plus = ...19)
#str(pop_age)

pop_age <- pop_age[-1,c(1,5,6,7,9,12:17,19)]
#str(pop_age)
 
pop_age <- bind_cols(pop_age[,1], lapply(pop_age[,c(2:12)], as.numeric))
str(pop_age)

# Save cleaned data
pop_age %>%
  write.csv("data_cleaned/db_pop_age.csv",row.names=FALSE)
```

## Population by sex

Source: IBGE, table 2093, https://sidra.ibge.gov.br/tabela/2093#:~:text=Tabela%202093%3A%20Popula%C3%A7%C3%A3o%20residente%20por,Amostra%20%2D%20Caracter%C3%ADsticas%20Gerais%20da%20Popula%C3%A7%C3%A3o

```{r}
# Population by sex
pop_sex <- read_xlsx("data/IBGE/Censo_tabela2093_pop_sexo.xlsx", sheet = 1) #local path of database
#str(pop_sex)
a <- dim(pop_sex)

pop_sex <- pop_sex[-c(1,2,3,4,5,6,7,a[1]),]

pop_sex <- pop_sex %>%
  rename(city_UF = "Tabela 2093 - População residente por cor ou raça, sexo, situação do domicílio e grupos de idade - Amostra - Características Gerais da População",
         pop.male = ...4,
         pop.female = ...5)
#str(pop_sex)

pop_sex <- pop_sex[,c(1,4,5)]
#str(pop_sex)

pop_sex <- bind_cols(pop_sex[,1], lapply(pop_sex[,c(2,3)], as.numeric))
str(pop_sex)

# Save cleaned data
pop_sex %>%
  write.csv("data_cleaned/db_pop_sex.csv",row.names=FALSE)
```

## Population rural vs urban

Source: IBGE, table 2093, https://sidra.ibge.gov.br/tabela/2093#:~:text=Tabela%202093%3A%20Popula%C3%A7%C3%A3o%20residente%20por,Amostra%20%2D%20Caracter%C3%ADsticas%20Gerais%20da%20Popula%C3%A7%C3%A3o

```{r}
# Population rural vs urban
pop_urban <- read_xlsx("data/IBGE/Censo_tabela2093_pop_ruralVSurbana.xlsx", sheet = 1) #local path of database
#str(pop_urban)
a <- dim(pop_urban)

pop_urban <- pop_urban[-c(1,2,3,4,5,6,7,8,a[1]),]

pop_urban <- pop_urban %>%
  rename(city_UF = "Tabela 2093 - População residente por cor ou raça, sexo, situação do domicílio e grupos de idade - Amostra - Características Gerais da População",
         pop.urban = ...3,
         pop.rural = ...4)
#str(pop_urban)
 
pop_urban <- pop_urban[,c(1,3,4)]
#str(pop_urban)

pop_urban <- bind_cols(pop_urban[,1], lapply(pop_urban[,c(2,3)], as.numeric))
str(pop_urban)

# Save cleaned data
pop_urban %>%
  write.csv("data_cleaned/db_pop_urban.csv",row.names=FALSE)
```

## Population by race

Source: IBGE, table 2093, https://sidra.ibge.gov.br/tabela/2093#:~:text=Tabela%202093%3A%20Popula%C3%A7%C3%A3o%20residente%20por,Amostra%20%2D%20Caracter%C3%ADsticas%20Gerais%20da%20Popula%C3%A7%C3%A3o

```{r}
# Population by race
pop_race <- read_xlsx("data/IBGE/Censo_tabela136_pop_raca.xlsx", sheet = 1) #local path of database
#str(pop_race)
a <- dim(pop_race)

pop_race <- pop_race[-c(1,2,3,a[1]),]

pop_race <- pop_race %>%
  rename(city_UF = "Tabela 136 - População residente, por cor ou raça",
         pop.white = ...17,
         pop.black = ...18,
         pop.asian = ...19,
         pop.pardos = ...20,
         pop.native = ...21,
         pop.others = ...22)
#str(pop_race)
  
pop_race <- pop_race[-c(1,2),c(1,17,18,19,20,21,22)]
#str(pop_race)

pop_race[ pop_race == "-" ] <- "0"
#str(pop_race)

pop_race <- bind_cols(pop_race[,1], lapply(pop_race[,c(2:7)], as.numeric))
str(pop_race)

# Save cleaned data
pop_race %>%
  write.csv("data_cleaned/db_pop_race.csv",row.names=FALSE)
```

## Population by education

Source: IBGE, table 1972, https://sidra.ibge.gov.br/tabela/1972

```{r}
# Population by education level
pop_edu.lvl <- read_xlsx("data/IBGE/Censo_tabela1972_pop_educacaototal.xlsx", sheet = 1) #local path of database
#str(pop_edu.lvl)
a <- dim(pop_edu.lvl)

pop_edu.lvl <- pop_edu.lvl[-c(1,2,3,4,5,6,a[1]),]
 
pop_edu.lvl <- pop_edu.lvl %>%
  rename(city_UF = "Tabela 1972 - Pessoas que frequentavam creche ou escola por nível e rede de ensino",
         pop.edu.kindergarden = ...3, #PT: creche 
         pop.edu.pre = ...4, #PT: pré-escolar
         pop.edu.literacy = ...5, #PT: alfabetização
         pop.edu.literacy_adults = ...6, #PT: alfabetização de jovens e adultos
         pop.edu.elem = ...7, #PT: fundamental
         pop.edu.elem_adults = ...8, #PT: fundamental de jovens e adultos
         pop.edu.high = ...9, #PT: ensino médio
         pop.edu.high_adults = ...10, #PT: ensino médio de jovens e adultos
         pop.edu.undergrad = ...11, #PT: graduação
         pop.edu.grad = ...12, #PT: especialização
         pop.edu.master = ...13, #PT: mestrado
         pop.edu.doc = ...14) #PT: doutorado
#str(pop_edu.lvl)

pop_edu.lvl <- pop_edu.lvl[,-c(2)]
#str(pop_edu.lvl)
 
pop_edu.lvl[ pop_edu.lvl == "-" ] <- "0"
#str(pop_edu.lvl)
 
pop_edu.lvl <- bind_cols(pop_edu.lvl[,1], lapply(pop_edu.lvl[,c(2:13)], as.numeric))
str(pop_edu.lvl)
```

```{r}
# Population by education type (private vs public)
pop_edu.type <- read_xlsx("data/IBGE/Censo_tabela1972_pop_educacaoprivadaVSpublica.xlsx", sheet = 1) #local path of database
#str(pop_edu.type)
a <- dim(pop_edu.type)

pop_edu.type <- pop_edu.type[-c(1:6,a[1]),]
 
pop_edu.type <- pop_edu.type %>%
  rename(city_UF = "Tabela 1972 - Pessoas que frequentavam creche ou escola por nível e rede de ensino",
         pop.edu.public = ...3,
         pop.edu.private = ...4)
#str(pop_edu.type)
 
pop_edu.type <- pop_edu.type[,-c(2)]
#str(pop_edu.type)
  
pop_edu.type[ pop_edu.type == "-" ] <- "0"
#str(pop_edu.type)
  
pop_edu.type <- bind_cols(pop_edu.type[,1], lapply(pop_edu.type[,c(2:3)], as.numeric))
str(pop_edu.type)

#Join education levels and type data sets
pop_edu <- left_join(pop_edu.lvl, pop_edu.type, by = "city_UF")
str(pop_edu)

# Match with city codes
# WIP  


# Save cleaned data
# pop_edu %>%
#   write.csv("data_cleaned/db_pop_edu.csv",row.names=FALSE)
```


# Covid data

## Deaths and cases

Source: https://brasil.io/dataset/covid19/caso_full/


```{r}
covid <- read.csv("data/Brasil.io/caso_full.csv") #local path of database

covid$date <- as.Date(covid$date, "%Y-%m-%d")
str(covid) 
```

```{r find last day of complete data for all cities, warning=FALSE}

last <- covid %>% 
  filter(place_type == "city") %>%
    filter(is_last == "True" ) %>%
      group_by(city_ibge_code) %>%
      summarise(last_day_available = date) 

last_day <- min(last$last_day_available)
```

```{r calculate cumulate cases and deaths until last day of available date}

db_covid_last <- covid %>% 
 filter(place_type == "city") %>%
  filter(city_ibge_code != "") %>% 
    filter(date <= last_day) %>%
       group_by(city_ibge_code) %>%
       summarise(cases = sum(new_confirmed), deaths = sum(new_deaths)) %>%
       arrange(city_ibge_code)

sum(is.na(db_covid_last)) #check if there is any missing cities

# Save cleaned data
db_covid_last %>%
  write.csv("data_cleaned/db_Covid_cases&deaths.csv",row.names=FALSE)
```

## Dedicated ICU beds sponsored by the Federal Government

Source: Localiza SUS, https://qsprod.saude.gov.br/extensions/DEMAS_C19Insumos_LEITOS_2021UTI/DEMAS_C19Insumos_LEITOS_2021UTI.html

```{r}
fed_covid_ICUbeds <- read.csv("data/Localiza_SUS/MS_Covid19_Leitos2021_19-04-21.csv", encoding = "UTF-8") #local path of database
#str(fed_covid_ICUbeds)

fed_covid_ICUbeds <- fed_covid_ICUbeds %>%
  select(Município, UF, Ano, Autorizações.publicadas, Valor.autorizações) %>%
  rename(PT_city = Município, Year = Ano, beds_authorized = Autorizações.publicadas, beds_authorized_cost = Valor.autorizações)
str(fed_covid_ICUbeds)
```

```{r}
aux_fed_covid_ICUbeds <- fed_covid_ICUbeds %>%
  filter(Year == 2021)%>%
    mutate(city_UF = paste0(PT_city," (",UF,")")) %>%
      group_by(city_UF)%>%
      summarise(beds = sum(beds_authorized), beds_cost = sum(beds_authorized_cost))

# anti_join(aux_fed_covid_ICUbeds, names_aux, by = c("city_UF" = "PT_city_UF_aux"))

db_fed_covid_ICUbeds <- full_join(aux_fed_covid_ICUbeds, names_aux, by = c("city_UF" = "PT_city_UF_aux")) %>%
  group_by(city_code)%>%
    summarise(beds = sum(beds), beds_cost = sum(beds_cost))

sum(is.na(db_fed_covid_ICUbeds$city_code)) #check if there is any missing or in excess cities

# Save cleaned data
db_fed_covid_ICUbeds %>%
  write.csv("data_cleaned/db_fed_covid_ICUbeds.csv",row.names=FALSE)

```

## Government expenditure with COVID

Source: Localiza SUS,  https://qsprod.saude.gov.br/extensions/DEMAS_C19SIOPS/DEMAS_C19SIOPS.html

```{r}
covid_spent <- read.csv("data/Localiza_SUS/MS_Covid19_Recursos_v2.csv", encoding = "UTF-8") #local path of database
#str(fed_covid_spent)

covid_spent <- covid_spent %>%
  mutate(city_UF = paste0(Município," (",UF,")")) %>%
  select(city_UF, Origem, Ano, Período, Valor.Empenhado..R...1, Valor.Pago..R...1) %>%
  rename(year = Ano, source = Origem, period = Período, 
         amount_committed = Valor.Empenhado..R...1, amount_disbursed = Valor.Pago..R...1)
str(covid_spent)
```
```{r}
aux_covid_spent <- covid_spent %>%
  filter(year == "2020") %>%
  group_by(city_UF, source, period) %>%
  summarise(amount_committed=sum(amount_committed), amount_disbursed=sum(amount_disbursed)) %>%
    pivot_wider(names_from = period, values_from = c(amount_committed, amount_disbursed)) %>%
      rename(exp_committed_p2 = "amount_committed_2º Bimestre",
             exp_committed_p3 = "amount_committed_3º Bimestre",
             exp_committed_p4 = "amount_committed_4º Bimestre",
             exp_committed_p5 = "amount_committed_5º Bimestre",
             exp_committed_p6 = "amount_committed_6º Bimestre",
             exp_disbursed_p2 = "amount_disbursed_2º Bimestre",
             exp_disbursed_p3 = "amount_disbursed_3º Bimestre",
             exp_disbursed_p4 = "amount_disbursed_4º Bimestre",
             exp_disbursed_p5 = "amount_disbursed_5º Bimestre",
             exp_disbursed_p6 = "amount_disbursed_6º Bimestre") %>%
      pivot_wider(names_from = source, 
                  values_from = c(exp_committed_p2, exp_committed_p3, exp_committed_p4, exp_committed_p5, exp_committed_p6,
                                  exp_disbursed_p2, exp_disbursed_p3, exp_disbursed_p4, exp_disbursed_p5, exp_disbursed_p6)) %>%
        rename(exp_committed_p2_city = "exp_committed_p2_Recurso Próprio",
               exp_committed_p3_city = "exp_committed_p3_Recurso Próprio",
               exp_committed_p4_city = "exp_committed_p4_Recurso Próprio",
               exp_committed_p5_city = "exp_committed_p5_Recurso Próprio",
               exp_committed_p6_city = "exp_committed_p6_Recurso Próprio",
               exp_disbursed_p2_city = "exp_disbursed_p2_Recurso Próprio",
               exp_disbursed_p3_city = "exp_disbursed_p3_Recurso Próprio",
               exp_disbursed_p4_city = "exp_disbursed_p4_Recurso Próprio",
               exp_disbursed_p5_city = "exp_disbursed_p5_Recurso Próprio",
               exp_disbursed_p6_city = "exp_disbursed_p6_Recurso Próprio",
               exp_committed_p2_state = "exp_committed_p2_Repasse Estadual",
               exp_committed_p3_state = "exp_committed_p3_Repasse Estadual",
               exp_committed_p4_state = "exp_committed_p4_Repasse Estadual",
               exp_committed_p5_state = "exp_committed_p5_Repasse Estadual",
               exp_committed_p6_state = "exp_committed_p6_Repasse Estadual",
               exp_disbursed_p2_state = "exp_disbursed_p2_Repasse Estadual",
               exp_disbursed_p3_state = "exp_disbursed_p3_Repasse Estadual",
               exp_disbursed_p4_state = "exp_disbursed_p4_Repasse Estadual",
               exp_disbursed_p5_state = "exp_disbursed_p5_Repasse Estadual",
               exp_disbursed_p6_state = "exp_disbursed_p6_Repasse Estadual",
               exp_committed_p2_fed = "exp_committed_p2_Repasse Federal",
               exp_committed_p3_fed = "exp_committed_p3_Repasse Federal",
               exp_committed_p4_fed = "exp_committed_p4_Repasse Federal",
               exp_committed_p5_fed = "exp_committed_p5_Repasse Federal",
               exp_committed_p6_fed = "exp_committed_p6_Repasse Federal",
               exp_disbursed_p2_fed = "exp_disbursed_p2_Repasse Federal",
               exp_disbursed_p3_fed = "exp_disbursed_p3_Repasse Federal",
               exp_disbursed_p4_fed = "exp_disbursed_p4_Repasse Federal",
               exp_disbursed_p5_fed = "exp_disbursed_p5_Repasse Federal",
               exp_disbursed_p6_fed = "exp_disbursed_p6_Repasse Federal")
# names(aux_covid_spent)

# match city with codes

aux2_covid_spent <- right_join(aux_covid_spent, names_aux2, by = c("city_UF" = "PT_city_UF_aux")) %>%
  group_by(city_code) %>%
   summarise(exp_committed_p2_city = sum(exp_committed_p2_city),
             exp_committed_p3_city = sum(exp_committed_p3_city),
             exp_committed_p4_city = sum(exp_committed_p4_city),
             exp_committed_p5_city = sum(exp_committed_p5_city),
             exp_committed_p6_city = sum(exp_committed_p6_city),
             exp_disbursed_p2_city = sum(exp_disbursed_p2_city),
             exp_disbursed_p3_city = sum(exp_disbursed_p3_city),
             exp_disbursed_p4_city = sum(exp_disbursed_p4_city),
             exp_disbursed_p5_city = sum(exp_disbursed_p5_city),
             exp_disbursed_p6_city = sum(exp_disbursed_p6_city),
             exp_committed_p2_state = sum(exp_committed_p2_state),
             exp_committed_p3_state = sum(exp_committed_p3_state),
             exp_committed_p4_state = sum(exp_committed_p4_state),
             exp_committed_p5_state = sum(exp_committed_p5_state),
             exp_committed_p6_state = sum(exp_committed_p6_state),
             exp_disbursed_p2_state = sum(exp_disbursed_p2_state),
             exp_disbursed_p3_state = sum(exp_disbursed_p3_state),
             exp_disbursed_p4_state = sum(exp_disbursed_p4_state),
             exp_disbursed_p5_state = sum(exp_disbursed_p5_state),
             exp_disbursed_p6_state = sum(exp_disbursed_p6_state),
             exp_committed_p2_fed = sum(exp_committed_p2_fed),
             exp_committed_p3_fed = sum(exp_committed_p3_fed),
             exp_committed_p4_fed = sum(exp_committed_p4_fed),
             exp_committed_p5_fed = sum(exp_committed_p5_fed),
             exp_committed_p6_fed = sum(exp_committed_p6_fed),
             exp_disbursed_p2_fed = sum(exp_disbursed_p2_fed),
             exp_disbursed_p3_fed = sum(exp_disbursed_p3_fed),
             exp_disbursed_p4_fed = sum(exp_disbursed_p4_fed),
             exp_disbursed_p5_fed = sum(exp_disbursed_p5_fed),
             exp_disbursed_p6_fed = sum(exp_disbursed_p6_fed))

sum(is.na(aux2_covid_spent$city_code)) #check if there is any missing cities

db_covid_spent <- aux2_covid_spent %>%
  filter(city_code != "")

# Save cleaned data
db_covid_spent %>%
  write.csv("data_cleaned/db_covid_spent.csv",row.names=FALSE)

```
          
## Vaccines

Source: Localiza SUS, https://qsprod.saude.gov.br/extensions/DEMAS_C19Vacina/DEMAS_C19Vacina.html


```{r}
vaccines <- read.csv("data/Localiza_SUS/Vacinas aplicadas_19-04-21.csv", encoding = "UTF-8") #local path of database
str(vaccines)

vaccines  <- vaccines %>%
  rename(city = X.U.FEFF..U.FEFF.Município) %>%
  mutate(city_UF = paste0(city," (",UF,")")) %>%
  select(city, UF, city_UF, Fabricante, Doses.Aplicadas, Dose.1, Dose.2) %>%
  rename(vac_provider = Fabricante, vac_shots = Doses.Aplicadas, vac_dose1 = Dose.1, vac_dose2 = Dose.2) %>%
  mutate(vac_provider = ifelse(vac_provider == "Vacina Adsorvida Covid-19 (Inativada) - Coronavac", "Coronavac", 
                                   ifelse(vac_provider == "Vacina Covid-19 - Covishield", "Covishield.AZ", "Unknown")))
str(vaccines)
```

```{r}
vaccines_aux <- vaccines %>%
  filter(UF != "-" & city != "-") %>%
    select(city_UF, vac_provider, vac_shots, vac_dose1, vac_dose2) %>%
        group_by(city_UF, vac_provider) %>%
        summarise(vac_shots = sum(vac_shots), vac_dose1 = sum(vac_dose1), vac_dose2 = sum(vac_dose2)) %>%
          pivot_wider(names_from = vac_provider, values_from = c(vac_shots, vac_dose1, vac_dose2)) %>%
            right_join(names, by = c("city_UF" = "PT_city_UF")) %>%
              group_by(city_code) %>%
                select(city_code, 
                    vac_shots_Coronavac, vac_shots_Covishield.AZ, vac_shots_Unknown,
                    vac_dose1_Coronavac, vac_dose1_Covishield.AZ, vac_dose1_Unknown, 
                    vac_dose2_Coronavac, vac_dose2_Covishield.AZ, vac_dose2_Unknown) %>%
                  arrange(city_code)

# Save cleaned data
vaccines_aux %>%
  write.csv("data_cleaned/db_vaccines.csv",row.names=FALSE)

```
## Federal covid emergency aid

Source: VIS DATA 3 beta, Auxílio Emergencial,  

https://aplicacoes.mds.gov.br/sagi/vis/data3/v.php?q[]=oNOtlcPavaarrLFrsWvJf7Od086vnG1ljah9aGl4ZnWraX9%2FZF2KjmObua5%2Bo7CjnbSardyedY6tYIyNY5NolszcuZmvq5piua%2BCgWNdgIyW2Z6fwu6wmWSIq26tcHl%2BYFi60JTWoqbA4HV0sntrbqVpdndYqpLHlNawmJjvv6mhg3Cg2qXAs3JoytagjbhUwOquoKG7mJ%2BhecN%2Fb1%2BDkVyVoKK%2B57Knn61deu9shYdjXYCMltmen8LusJlkiKtzr2V9d2KQxsKfz7CWwqONqm9%2BamapYnixpo7DxqbNolud8YBrbXRlY6ScvK%2BjksrEmJJ9qZCyhGBscWCd6Jq5s6qQvIlz4HBrkKd9XWerpJvlnsCxnFWX12aac1%2BNpHiXq6mhn%2BycsnZ3o4yVX5pmXsDqrqChu5ifoXnDf2hhg5FclaCivueyp5%2BtXXrvan2AY12AjJbZnp%2FC7rCZZIira6tveX5gWLrQlNaipsDgdXSyeWpupWl2eZqcuM2Y3aCYhbvDZnN4YWqiZLC9mJm81JbPZXPTrYVmaHheZdyorrqcoLrGW6qzZZaveWRlc5ip2qWywZqSf6GpnW5liat2X5%2B3lqberLCzX23NlGmWbVyI3ryVqK2ond5hjcRoYY%2BNY5NolszcuZmvq5piua9%2Bfm9Zh4pezayUyeDAl6FwdbCqa316Z1aCxKLLqZjQ3rJcfL5ncatlfXdikMbCn8%2BwlsKjjapugG1mqWJ4saaOw8amzaJbnfGAZGx0ZWOknLyvo5LKxJiSfamOr31gbHFgneiaubOqkLyJc%2BBuapanfV1nq6Sb5Z7AsZxVl9dkoXBfjaR4l6upoZ%2FsnLJ2d6OImGmWbVyI3ryVqK2ond5hjcRqY4eNY5NolszcuZmvq5piua%2BBgW9Zh4pezayUyeDAl6FwdbCtboJ6Z1aCxKLLqZjQ3rJcfL5pb7FlfXdikMbCn8%2BwlsKjjap1eWxmqWJ4saaOw8amzaJbnfGGZm50ZWOknLyvo5LKxJiSfamWs4BgbHFgneiaubOqkLyJc%2BB2a5OnfV1dxXCg2qXAs3KhydaYpXiZvufAmXeDqK%2FmXMOBbGi9wp%2Fdom7R7cKZd4Obm%2BWssolyoMzOVuVelszcuZmvq5piua%2BAf2lZh4pezayUyeDAl6FwdbCsb3l%2BYFi60JTWoqbA4HV0snlpcqVpdnmanLjNmN2gmIW7w21ydGVjpJy8r6OSysSYkn2pjquFYGxxYJ3omrmzqpC8iXPgbmWNp31dZ6ukm%2BWewLGcVZfXZaFvX42keJerqaGf7JyydnejiZlrlm1ciN68laitqJ3eYY3Eal2HjWOTXrCY4a6gr61wruuusolyk7jNps94btDwuleyfWh135q5wZxoy9Ooz3huw9y5p6GDcK3upnDJWJDGwp%2FPsJbCo42qb3hrZqlieLGmjsPGps2iW53xgmhoeF5l3KiuupygusZbqrNkjq95ZGVzmKnapbLBmpJ%2FoambbWWJq3Zfn7eWpt6ssLNfbc2SZaBpY4amsKOdtJqt3J51jq1ejJVfmmZewOquoKG7mJ%2BhecOAbl2DkVyVoKK%2B57Knn61deu9rhYBjXYCMltmen8LusJlkiKtssm15fmBO1JyZy6mmwrbBprGtcHXfmrnBnGiS1KjXYKmQsYFvoqmhrd50wcCskpKcmcuppsK2iKextVi1mpy8r6OSysSYkn2pkLOGYGxxYJ3omrmzqpC8iXPgcGmSp31dZ6ukm%2BWewLGcVZfXZqFuX42keJerqaGf7Jyydnejiphqlm1ciN68laitqJ3eYY3EamWKjWOTXrCY4a6gr61wruuusolyk7jNps94btDwuleyfGlxtJ%2BuuqqSktWl36JumOGuoK%2BtcHXsrrpxsk660JTWoqbA4HV0snlpaqVpdnmanLjNmN2gmIW7w2VzgWFqomSwvZiZvNSWz2Vz06yEZ2h4XmXcqK66nKC6xluqs2SUsXlkZXOYqdqlssGakn%2BhqZt1ZYmrdl%2Bft5am3qyws19tzZRpmmljhqawo520mq3cnnWOrWGKmV%2BaZl7A6q6gobuYn6F5w4JsYoORXJWgor7nsqefrV16722ChmNdgIyW2Z6fwu6wmWSIq3OqcHl%2BYFi60JTWoqbA4HV0soFnbKVpdnmanLjNmN2gmIW7w210e2FqomSwvZiZvNSWz2Vz07SFamh4Xlv2dLOvo6C8nKfcspiYtrOVqLuadbSswruzfbzUptmepn3guZmjC%2BKw3qLAbpicd6Ko4gDgyeS8VIG1mqzgnruxoI7DhInLqaLPm8GjsKmhWtpZwLOpTcnGo8uwpr7fvFSsraGpmXrCxvraw8qiioKgwu20maqrnpvlXJ2zqqDGwqaKop%2FC4hDhsq2erZmdvG6n8BHDn9Ogon2%2Brpidu6ms6FkQ6KWWutBTkrCYypuvo6i7lmOcj666pp931aLenp993G2nobpVrN6prsGqjrvQU9qepb6bvff2qqGj3KhtkZiRuNSn3KxTIBW7nZ%2B3VWLsnrpumZzD1JSTYIPC7sCjnbtVn%2BWetBHko7zKpoqhon3rEO6etJ6d6FmOvqOWusKn07Oifb6unbSpWJDapbzAV6HG1ZTWXZR97rKmXLqaqtqswK%2BbnHfRlNyeU80%2BB5aosZipmXq9uqCQuNWc4KxToNy2rJ1rhZ%2FsrLyvqk28zZjRAODT4LanXLikrJmmsremTaHWl9OgnL7ncIqdtKSsma28wpiZd8JT3aKlfe2ypJ27qJvdqG2%2BmJ%2B4gaMt95XJ5LCjXJKqnuKctq%2BjUKfGpt2slNCbsqChr%2Fjn7562wVeRxoGjLfeVyeSwo1yKpKbsmm2UmJoaDp%2FTnlaz3Lmjrmipqe2auW6YTcrGpYqvmM3cwKedrKRa6Zq%2Fr1edGhuV1qaWzJuPo6i7llq%2FmroR5JnAwq%2FaeA%3D%3D


```{r}
emergency_aid <- read.csv("data/VIS DATA 3/Auxilio Emergencial_2020-21.csv") #local path of database
str(emergency_aid)

emergency_aid <- emergency_aid %>%
  mutate(city_UF = paste0(Unidade.Territorial," (",UF,")")) %>%
  rename(year = Referência, 
         aid_people_CU = Pessoas.elegíveis.do.público.Cadastro.Único..sem.bolsa., 
         aid_value_CU = Valor.total.a.ser.repassado.para.público.Cadastro.Único..sem.bolsa., 
         aid_people_appCaixa = Pessoas.elegíveis.do.público.Aplicativo.Caixa, 
         aid_value_appCaixa = Valor.total.a.ser.repassado.para.público.Aplicativo.Caixa, 
         aid_people_lawsuit = Pessoas.elegíveis.por.meio.Judicial, 
         aid_value_lawsuit = Valor.total.a.ser.repassado.para.público.Judicial,
         aid_people_BolsaFamilia = Pessoas.elegíveis.do.público.Bolsa.Família,
         aid_value_BolsaFamilia = Valor.total.a.ser.repassado.para.público.Bolsa.Família,
         aid_people = Total.de.pessoas.Elegíveis) %>%
  mutate(aid_value = aid_value_CU + aid_value_appCaixa + aid_value_lawsuit + aid_value_BolsaFamilia) %>%
  select(-Código, -Unidade.Territorial, -UF)
  
```

```{r}
emergency_aid_aux <- emergency_aid %>%
  filter(year == 2020) %>%
  group_by(city_UF) %>%
    summarise(aid_people_CU = sum(aid_people_CU), 
              aid_people_appCaixa = sum(aid_people_appCaixa), 
              aid_people_lawsuit = sum(aid_people_lawsuit), 
              aid_people_BolsaFamilia = sum(aid_people_BolsaFamilia),
              aid_people = sum(aid_people),
              aid_value_CU = sum(aid_value_CU), 
              aid_value_appCaixa = sum(aid_value_appCaixa), 
              aid_value_lawsuit = sum(aid_value_lawsuit), 
              aid_value_BolsaFamilia = sum(aid_value_BolsaFamilia),
              aid_value = sum(aid_value))
str(emergency_aid_aux)

db_emergency_aid <- emergency_aid_aux %>%
  left_join(names_aux3, by = c("city_UF" = "PT_city_UF_aux")) %>%
  select(-city_UF) %>%
    arrange(city_code)
    
db_emergency_aid <- db_emergency_aid[, c(11, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)] 


# Save cleaned data
db_emergency_aid %>%
  write.csv("data_cleaned/db_emergency_aid.csv",row.names=FALSE)

```

# Economic data

## GDP by city (2010-18)

```{r}
# Total GDP
GDP.2010_14 <- read_xlsx("data/IBGE/tabela5938_PIB_2010-14.xlsx", sheet = 1) #local path of database
str(GDP.2010_14)
a <- dim(GDP.2010_14)

GDP.2010_14 <- GDP.2010_14[-c(1,2,3,a[1]),]
  
GDP.2010_14 <- GDP.2010_14 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.2010 = ...2,
         GDP.2011 = ...3,
         GDP.2012 = ...4,
         GDP.2013 = ...5,
         GDP.2014 = ...6)


GDP.2015_18 <- read_xlsx("data/IBGE/tabela5938_PIB_2015-18.xlsx", sheet = 1) #local path of database
str(GDP.2015_18)
a <- dim(GDP.2015_18)

GDP.2015_18 <- GDP.2015_18[-c(1,2,3,a[1]),]
  
GDP.2015_18 <- GDP.2015_18 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.2015 = ...2,
         GDP.2016 = ...3,
         GDP.2017 = ...4,
         GDP.2018 = ...5)
```

```{r}
# Tax GDP
GDP.tax.2010_14 <- read_xlsx("data/IBGE/tabela5938_PIB_2010-14.xlsx", sheet = 2) #local path of database
str(GDP.tax.2010_14)
a <- dim(GDP.tax.2010_14)

GDP.tax.2010_14 <- GDP.tax.2010_14[-c(1,2,3,a[1]),]
  
GDP.tax.2010_14 <- GDP.tax.2010_14 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.tax.2010 = ...2,
         GDP.tax.2011 = ...3,
         GDP.tax.2012 = ...4,
         GDP.tax.2013 = ...5,
         GDP.tax.2014 = ...6)


GDP.tax.2015_18 <- read_xlsx("data/IBGE/tabela5938_PIB_2015-18.xlsx", sheet = 2) #local path of database
str(GDP.tax.2015_18)
a <- dim(GDP.tax.2015_18)

GDP.tax.2015_18 <- GDP.tax.2015_18[-c(1,2,3,a[1]),]
  
GDP.tax.2015_18 <- GDP.tax.2015_18 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.tax.2015 = ...2,
         GDP.tax.2016 = ...3,
         GDP.tax.2017 = ...4,
         GDP.tax.2018 = ...5)
```  

```{r}
# Agriculture GDP
GDP.agri.2010_14 <- read_xlsx("data/IBGE/tabela5938_PIB_2010-14.xlsx", sheet = 4) #local path of database
str(GDP.agri.2010_14)
a <- dim(GDP.agri.2010_14)

GDP.agri.2010_14 <- GDP.agri.2010_14[-c(1,2,3,a[1]),]
  
GDP.agri.2010_14 <- GDP.agri.2010_14 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.agri.2010 = ...2,
         GDP.agri.2011 = ...3,
         GDP.agri.2012 = ...4,
         GDP.agri.2013 = ...5,
         GDP.agri.2014 = ...6)

GDP.agri.2015_18 <- read_xlsx("data/IBGE/tabela5938_PIB_2015-18.xlsx", sheet = 4) #local path of database
str(GDP.agri.2015_18)
a <- dim(GDP.agri.2015_18)

GDP.agri.2015_18 <- GDP.agri.2015_18[-c(1,2,3,a[1]),]
  
GDP.agri.2015_18 <- GDP.agri.2015_18 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.agri.2015 = ...2,
         GDP.agri.2016 = ...3,
         GDP.agri.2017 = ...4,
         GDP.agri.2018 = ...5)
```

```{r}
# Industry GDP
GDP.industry.2010_14 <- read_xlsx("data/IBGE/tabela5938_PIB_2010-14.xlsx", sheet = 5) #local path of database
str(GDP.industry.2010_14)
a <- dim(GDP.industry.2010_14)

GDP.industry.2010_14 <- GDP.industry.2010_14[-c(1,2,3,a[1]),]
  
GDP.industry.2010_14 <- GDP.industry.2010_14 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.industry.2010 = ...2,
         GDP.industry.2011 = ...3,
         GDP.industry.2012 = ...4,
         GDP.industry.2013 = ...5,
         GDP.industry.2014 = ...6)


GDP.industry.2015_18 <- read_xlsx("data/IBGE/tabela5938_PIB_2015-18.xlsx", sheet = 5) #local path of database
str(GDP.industry.2015_18)
a <- dim(GDP.industry.2015_18)

GDP.industry.2015_18 <- GDP.industry.2015_18[-c(1,2,3,a[1]),]
  
GDP.industry.2015_18 <- GDP.industry.2015_18 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.industry.2015 = ...2,
         GDP.industry.2016 = ...3,
         GDP.industry.2017 = ...4,
         GDP.industry.2018 = ...5)

```

```{r}
# Services GDP
GDP.services.2010_14 <- read_xlsx("data/IBGE/tabela5938_PIB_2010-14.xlsx", sheet = 6) #local path of database
str(GDP.services.2010_14)
a <- dim(GDP.services.2010_14)

GDP.services.2010_14 <- GDP.services.2010_14[-c(1,2,3,a[1]),]
  
GDP.services.2010_14 <- GDP.services.2010_14 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.services.2010 = ...2,
         GDP.services.2011 = ...3,
         GDP.services.2012 = ...4,
         GDP.services.2013 = ...5,
         GDP.services.2014 = ...6)


GDP.services.2015_18 <- read_xlsx("data/IBGE/tabela5938_PIB_2015-18.xlsx", sheet = 6) #local path of database
str(GDP.services.2015_18)
a <- dim(GDP.services.2015_18)

GDP.services.2015_18 <- GDP.services.2015_18[-c(1,2,3,a[1]),]
  
GDP.services.2015_18 <- GDP.services.2015_18 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.services.2015 = ...2,
         GDP.services.2016 = ...3,
         GDP.services.2017 = ...4,
         GDP.services.2018 = ...5)
```

```{r}
# Government GDP
GDP.govt.2010_14 <- read_xlsx("data/IBGE/tabela5938_PIB_2010-14.xlsx", sheet = 7) #local path of database
str(GDP.govt.2010_14)
a <- dim(GDP.govt.2010_14)

GDP.govt.2010_14 <- GDP.govt.2010_14[-c(1,2,3,a[1]),]
  
GDP.govt.2010_14 <- GDP.govt.2010_14 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.govt.2010 = ...2,
         GDP.govt.2011 = ...3,
         GDP.govt.2012 = ...4,
         GDP.govt.2013 = ...5,
         GDP.govt.2014 = ...6)


GDP.govt.2015_18 <- read_xlsx("data/IBGE/tabela5938_PIB_2015-18.xlsx", sheet = 7) #local path of database
str(GDP.govt.2015_18)
a <- dim(GDP.govt.2015_18)

GDP.govt.2015_18 <- GDP.govt.2015_18[-c(1,2,3,a[1]),]
  
GDP.govt.2015_18 <- GDP.govt.2015_18 %>%
  rename(city_UF = "Tabela 5938 - Produto interno bruto a preços correntes, impostos, líquidos de subsídios, sobre produtos a preços correntes e valor adicionado bruto a preços correntes total e por atividade econômica, e respectivas participações - Referência 2010",
         GDP.govt.2015 = ...2,
         GDP.govt.2016 = ...3,
         GDP.govt.2017 = ...4,
         GDP.govt.2018 = ...5)
```

```{r}
GDP_aux <- GDP.2015_18 %>%
  select(city_UF)

# Select GDP data with economy sector (2018)

GDP_aux <- bind_cols(GDP_aux, 
          as.numeric(unlist(GDP.agri.2015_18$GDP.agri.2018)), 
          as.numeric(unlist(GDP.govt.2015_18$GDP.govt.2018)), 
          as.numeric(unlist(GDP.industry.2015_18$GDP.industry.2018)), 
          as.numeric(unlist(GDP.services.2015_18$GDP.services.2018)), 
          as.numeric(unlist(GDP.tax.2015_18$GDP.tax.2018))) %>%
  rename(GDP.agri.2018 = ...2, 
         GDP.govt.2018 = ...3, 
         GDP.industry.2018 = ...4, 
         GDP.services.2018 = ...5, 
         GDP.tax.2018 = ...6)
#str(GDP_aux)

# Calculate GDP % per economy sector

GDP_aux <- GDP_aux %>%
    mutate(GDP.2018 = rowSums(.[2:6])) %>%
      mutate(GDP.agri.2018_pct = GDP.agri.2018 / GDP.2018,
             GDP.govt.2018_pct = GDP.govt.2018 / GDP.2018,
             GDP.industry.2018_pct = GDP.industry.2018 / GDP.2018,
             GDP.services.2018_pct = GDP.services.2018 / GDP.2018,
             GDP.tax.2018_pct = GDP.tax.2018 / GDP.2018)
     
GDP_aux <- GDP_aux[,c(1,8,9,10,11,12,7)]
str(GDP_aux)
```

```{r}
GDP_aux2 <- GDP.2015_18 %>%
  select(city_UF)

# Select GDP data to calculate growth (2010, 15, 18)

GDP_aux2 <-  GDP_aux2 %>%
  bind_cols(as.numeric(unlist(GDP.2010_14$GDP.2010)), 
            as.numeric(unlist(GDP.2015_18$GDP.2015)),
            as.numeric(unlist(GDP_aux$GDP.2018))) %>%
        rename(GDP.2010 = ...2,
               GDP.2015 = ...3,
               GDP.2018 = ...4)
#str(GDP_aux2)

# Calculate annual nominal growth

GDP_aux2 <- GDP_aux2 %>%
    mutate(GDP.growth2010_15 = (GDP.2015/GDP.2010)^(1/5)-1) %>%
    mutate(GDP.growth2015_18 = (GDP.2018/GDP.2015)^(1/3)-1)

GDP_aux2 <- GDP_aux2[,c(5,6)]
str(GDP_aux2)
```

```{r}
# Calculate GDP per capita in 2018
GDP_aux3 <- GDP.2015_18 %>%
  select(city_UF) %>%
    bind_cols(pop$pop.2018, GDP_aux$GDP.2018) %>%
    mutate(GDP.percapita.2018 = ...3 / ...2)
str(GDP_aux3)

GDP_aux3 <- GDP_aux3[,c(4)]
str(GDP_aux3)

# Bind all GDP tables 
GDP <- bind_cols(GDP_aux[,c(1:6)], GDP_aux2, GDP_aux3)

# Match with city codes
GDP <- left_join(GDP, names_aux4, by = c("city_UF" = "PT_city_UF_aux"))
GDP <- GDP[,c(10,2:9)]
str(GDP)

# Save cleaned data
GDP %>%
  write.csv("data_cleaned/db_GDP.csv",row.names=FALSE)
```

[end]
