---
title: "Final Project - Covid in Brazil and associated variables"
author: "Alexandre Ferraz, Isabella Amorim, Lu Lu"
date: 'May, 2021'
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen = 0, digits = 3)
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(bestglm, glmnet, leaps, car, tidyverse, pROC, caret) # add the packages needed
library(xtable)
library(dplyr)
library(readxl)
```

```{r, echo=FALSE}
#clear variables
rm(list=ls())
```

\pagebreak


# Executive summary

## Goal of the study

## Data

## Findings



\pagebreak

# Introduction

## The Covid problem in Brazil

Brazil confirmed its first case of Covid-19 in February 26,2020. As of now, almost 15 months after the outbreak, the country sees more 15M of cases and  more than 420,000 deaths. Since February, Brazil is the country with more daily new deaths per population and the third in terms of total cases, after the US and India.

[insert figure with evolution of deaths per population of selected countries]

[insert figure with evolution of total cases of selected countries]

What are the reasons for such disaster in Brazil?

## Structural facts about Brazil

Brazil is an emerging country with 211M (c.64% of US population), median age of 33.5 years (vs 38.4 in the US) and GDP per capita US$ 8,700 (c.13% of US), according to the 2019 numbers of the World Bank. The country is relatively young, populous and poor, compared to other nations which suffered greatly with Covid, so it is important to investigate what internal characteristics contributed for such devastating crisis.

From an internal perspective, the country is greatly diverse in terms of geography, population, economy, health infrastructure and political mindset.

**Geography**

[plot Brazil map with states and number of cities]

- The country is the 5th largest in the world
- It has 26 states and 1 Federal District (where the capital Brasília is the only city located), so there are 27 territories part of the Federation
- Brazil is divided in 5 macro geographic regions: Southeast, Northeast, South, North, and Midwest
- There are 5570 counties, unevenly distributed across states. Minas Gerais (MG) and São Paulo (SP) have has more counties, 853 and 645 respectively, while Amapá (AP) and Roraima (RR) have only 16 and 15.

**Population**

[plot Brazil map with states and population]

- The 211M inhabitants are concentrated on the coast, especially in regions Southeast (X% of total) and Northeast (X% of total)
- The state of São Paulo (SP) corresponds to X% of Brazil's population, while its capital, São Paulo county, has XXM inhabitant ((X% of total)
- North and Midwest have smaller populations despite the huge territories, leading to lower population densities
- There are 48.4% men and 51.6% woman
- 85% of the population live in urban areas
- Brazilian uses an ethnic and skin color self-declared methodology to classify races. There are approximately 46.8% pardos*, 42.7% white, 9.4% black, 1.1% Asian or Indigenous (native Amerindians)

*According to the Brazilian Institute of Geography and Statistics (IBGE), 'pardos' represent a diverse range of skin colors and ethnic backgrounds with a skin tone darker than white and lighter than black, also be used for people from Asia with darker skin tones or other ethnicities with the same color

**Economy**

[plot Brazil map with states and % of GDP]

- The 211M inhabitants are concentrated on the coast, especially in regions Southeast (X% of total) and Northeast (X% of total)
- The state of São Paulo (SP) corresponds to X% of Brazil's population, while its capital, São Paulo county, has XXM inhabitant ((X% of total)
- North and Midwest have smaller populations despite the huge territories, leading to lower population densities
- There are 48.4% men and 51.6% woman
- 85% of the population live in urban areas



## Covid response

## Goal of the study







**Objective: run regression analysis of covid cumulative cases ans deaths vs independent variables**

# Data preparation

## Inputs: Covid deaths and cases vs HEALTH variables

```{r}
# Read COVID data
covid_cumulative <- read.csv("data_cleaned/db_covid_cumulative.csv") %>%
  arrange(city_code)
str(covid_cumulative)

# Read HEALTH database
health <- read.csv("data_joined/db_health.csv") %>%
  arrange(city_code)
str(health)

# Read STATE variable
state <- read.csv("data_cleaned/db_geo_state.csv") %>%
  arrange(city_code)
state <- state %>%
  mutate(state_name = as.factor(state)) %>%
  select(-state)
str(state)

# Read city names (to identify cities)
city <- read.csv("data/city_names_v1.csv", encoding = "UTF-8") %>%
  rename(city_code = X.U.FEFF.city_code)

```

## Preparation: format databases to run the regressions

```{r}
# Create transformed (log) independent variables
covid_cumulative <- covid_cumulative %>%
  mutate(logdeaths_per100K = log(deaths_per100K+0.01)) %>%
  mutate(logcases_per100K = log(cases_per100K+1))
```


```{r}
## CASES

# Prepare CASES base for regression (full = might have NAs)
cases_health_full <- covid_cumulative %>%
  select(city_code, cases_per100K) %>%
    left_join(state, by = "city_code") %>%
      left_join(health, by = "city_code") %>%
        select(-city_code)
#str(cases_health_full)
sum(is.na(cases_health_full))

# Prepare CASES base for LASSO (without NAs)
cases_health.NA <- colSums(is.na(cases_health_full))
cases_health.NA.index <- which(cases_health.NA != 0, arr.ind= TRUE)
cases_health <- cases_health_full[,-cases_health.NA.index]
sum(is.na(cases_health))

# Prepare LOG CASES base for regression (full = might have NAs)
logcases_health_full <- covid_cumulative %>%
  select(city_code, logcases_per100K) %>%
    left_join(state, by = "city_code") %>%
      left_join(health, by = "city_code") %>%
        select(-city_code)
#str(logcases_health_full)
sum(is.na(logcases_health_full))

# Prepare LOG CASES base for LASSO (without NAs)
logcases_health.NA <- colSums(is.na(logcases_health_full))
logcases_health.NA.index <- which(logcases_health.NA != 0, arr.ind= TRUE)
logcases_health <- logcases_health_full[,-logcases_health.NA.index]
sum(is.na(logcases_health))

## DEATHS

# Prepare DEATHS base for regression
deaths_health_full <- covid_cumulative %>%
  select(city_code, deaths_per100K) %>%
    left_join(state, by = "city_code") %>%
      left_join(health, by = "city_code") %>%
        select(-city_code)
#str(deaths_health_full)
sum(is.na(deaths_health_full))

# Prepare DEATHS base for LASSO (without NAs)
deaths_health.NA <- colSums(is.na(deaths_health_full))
deaths_health.NA.index <- which(deaths_health.NA != 0, arr.ind= TRUE)
deaths_health <- deaths_health_full[,-deaths_health.NA.index]
sum(is.na(deaths_health))

# Prepare LOG DEATHS base for regression
logdeaths_health_full <- covid_cumulative %>%
  select(city_code, logdeaths_per100K) %>%
    left_join(state, by = "city_code") %>%
      left_join(health, by = "city_code") %>%
        select(-city_code)
#str(logdeaths_health_full)
sum(is.na(logdeaths_health_full))

# Prepare LOG DEATHS base for LASSO (without NAs)
logdeaths_health.NA <- colSums(is.na(logdeaths_health_full))
logdeaths_health.NA.index <- which(logdeaths_health.NA != 0, arr.ind= TRUE)
logdeaths_health <- logdeaths_health_full[,-logdeaths_health.NA.index]
sum(is.na(logdeaths_health))

```

# Regressions

## Deaths: linear model (all in)

```{r}
# Run deaths vs health regression (linear model)
deaths_health_fit0 <- lm(deaths_per100K~., deaths_health_full)
summary(deaths_health_fit0)

# Run log deaths vs health regression (linear model)
logdeaths_health_fit0 <- lm(logdeaths_per100K~., logdeaths_health_full)
summary(logdeaths_health_fit0)
```

## Cases: linear model (all in)

```{r}
# Run cases vs health regression (linear model)
cases_health_fit0 <- lm(cases_per100K~., cases_health_full)
summary(cases_health_fit0)

# Run log cases vs health regression (linear model)
logcases_health_fit0 <- lm(logcases_per100K~., logcases_health_full)
summary(logcases_health_fit0)

```

## Deaths: LASSO and fine-tuning

```{r}
## Run LASSO of DEATHS vs health variables

#Step 1: Prepare design matrix
Y_deaths <- deaths_health[,1]
X_health <- model.matrix(deaths_per100K~., data = deaths_health)

#Step 2: Find x's output from LASSO with min cross-validation error
set.seed(10)  # to control the ramdomness in K folds 

deaths_health_fit.cv <- cv.glmnet(X_health, Y_deaths, alpha=1, nfolds=10, intercept = T, 
                                  penalty.factor = c(rep(0,27), rep(1, ncol(X_health)-27))) 

plot(deaths_health_fit.cv)

health_coef.min <- coef(deaths_health_fit.cv, s="lambda.min")  #s=c("lambda.1se","lambda.min") or lambda value
health_coef.min <- health_coef.min[which(health_coef.min !=0),]   # get the non-zero coefficients
health_var.min <- rownames(as.matrix(health_coef.min))[-1] # output the names  dim(as.matrix(coef.min))

#health_coef.min
health_var.min

```

```{r}
## Screen variables to avoid colinearity
#which( var.min == "beds.priv.per100Khab.2021_specialty.total", arr.ind = TRUE)
health_var.min_final <- health_var.min
dim(as.matrix(health_var.min_final))

```

```{r}
## Fine tuning

# Calculate number of levels of forced variable (states)
forced_size <- deaths_health %>% 
  select(state_name) %>%
    distinct() %>%
      dim()
forced_size <- forced_size[1]

# Calculate number of variables chosen in LASSO
max_var<-dim(as.matrix(health_var.min_final))[1] 

deaths_health.sub <- deaths_health %>%
  select(state_name, health_var.min_final[forced_size:max_var], deaths_per100K)

X_health.sub <- model.matrix(deaths_per100K~., data = deaths_health.sub)

deaths_health.fine_tuning <- regsubsets(deaths_per100K ~., data = deaths_health.sub, 
                                        nvmax=max_var, force.in = c(1:(forced_size-1)), method="exhaustive", really.big = TRUE)
 
summary(deaths_health.fine_tuning)
plot(summary(deaths_health.fine_tuning)$cp) 
```

```{r}
# Select number of (not-forced) variables using Cp
deaths_health.final.summary<- summary(deaths_health.fine_tuning)
opt.size <- 15 #selected using elbow rule

# Select variables
deaths_health.final.var.1 <- deaths_health.final.summary$which
deaths_health.final.var <- colnames(deaths_health.final.var.1)[deaths_health.final.var.1[opt.size, ]][-1] #take the intercept put
deaths_health.final.var <- deaths_health.final.var[-c(1:(forced_size-1))]
deaths_health.final.var

# Final model
deaths_health.final.fit <- lm(deaths_per100K ~., 
                              deaths_health.sub[, c("deaths_per100K", "state_name", deaths_health.final.var)])
summary(deaths_health.final.fit)

plot(deaths_health.final.fit, 1)
plot(deaths_health.final.fit, 2)
```

```{r}
deaths_health.final.predicted <- predict(deaths_health.final.fit)
deaths_health.final.stdres <- rstandard(deaths_health.final.fit)

plot(Y_deaths,deaths_health.final.predicted)

# Save data
deaths_health_results <- data.frame(covid_cumulative[,c("city_code","deaths_per100K")], 
                                    deaths_health.final.predicted, deaths_health.final.stdres) 
deaths_health_results <- deaths_health_results %>%
  left_join(city, by = "city_code")

deaths_health_results <- deaths_health_results[,c(1,7,6,2,3,4)]
str(deaths_health_results)

deaths_health_results %>%
  write.csv("results/db_results_deaths_health.csv",row.names=FALSE)
```

```{r}
# Select variables MANUALLY

deaths_health_manual <- deaths_health %>% 
   select(deaths_per100K, 
          state_name,
          hosp_beds.2021.per100Khab,
          hosp_beds.growth.2020to21,
          beds.total.per100Khab.2021_ICU.adult1,
          beds.total.per100Khab.2021_specialty.total,
          #beds.total.growth20to21_ICU.adult1,
          beds.total.growth20to21_ICU.adult3,
          #beds.priv.growth20to21_ICU.adult1,
          beds.priv.growth20to21_ICU.adult3,  
          beds.publ.per100Khab.2021_specialty.total,
          #oxygen.usedper100K_03.2020,
          #infecto_all.2020.totalper100K,
          nurses_prof.2020.totalper100K,
          deaths_diabetes.2019per100K,
          deaths_circulatory.2019per100K,
          #deaths_hypertensive.2019per100K,
          deaths_respiratory.2019per100K,
          lives_03.20.per100k.total,
          lives_per100k.growth.total)
  
# Run deaths vs health regression (linear model) - fit MANUAL
deaths_health_fit1 <- lm(deaths_per100K~., deaths_health_manual)
summary(deaths_health_fit1)
  
```

## log Deaths: LASSO and fine-tuning

```{r}
## Run LASSO of log DEATHS vs health variables

#Step 1: Prepare design matrix
Y_logdeaths <- logdeaths_health[,1]
X_loghealth <- model.matrix(logdeaths_per100K~., data = logdeaths_health)

#Step 2: Find x's output from LASSO with min cross-validation error
set.seed(10)  # to control the ramdomness in K folds 

logdeaths_health_fit.cv <- cv.glmnet(X_loghealth, Y_logdeaths, alpha=1, nfolds=10, intercept = T, 
                                  penalty.factor = c(rep(0,27), rep(1, ncol(X_loghealth)-27))) 

plot(logdeaths_health_fit.cv)

loghealth.coef.min <- coef(logdeaths_health_fit.cv, s="lambda.min")  #s=c("lambda.1se","lambda.min") or lambda value
loghealth.coef.min <- loghealth.coef.min[which(loghealth.coef.min !=0),]   # get the non-zero coefficients
loghealth.var.min <- rownames(as.matrix(loghealth.coef.min))[-1] # output the names  dim(as.matrix(coef.min))

#loghealth.coef.min
loghealth.var.min

```

```{r}
## Screen variables to avoid colinearity
#which( var.min == "beds.priv.per100Khab.2021_specialty.total", arr.ind = TRUE)
loghealth.var.min_final <- loghealth.var.min
dim(as.matrix(loghealth.var.min_final))

```

```{r}
## Fine tuning

# Calculate number of levels of forced variable (states)
forced_size <- logdeaths_health %>% 
  select(state_name) %>%
    distinct() %>%
      dim()
forced_size <- forced_size[1]

# Calculate number of variables chosen in LASSO
max_var<-dim(as.matrix(loghealth.var.min_final))[1] 

logdeaths_health.sub <- logdeaths_health %>%
  select(state_name, loghealth.var.min_final[forced_size:max_var], logdeaths_per100K)

X_loghealth.sub <- model.matrix(logdeaths_per100K~., data = logdeaths_health.sub)

logdeaths_health.fine_tuning <- regsubsets(logdeaths_per100K ~., data = logdeaths_health.sub, 
                                        nvmax=max_var, force.in = c(1:(forced_size-1)), method="forward")
 
summary(logdeaths_health.fine_tuning)
plot(summary(logdeaths_health.fine_tuning)$cp) 
```

```{r}
# Select number of (not-forced) variables using Cp
logdeaths_health.final.summary<- summary(logdeaths_health.fine_tuning)
opt.size <- 20 #selected using elbow rule

# Select variables
logdeaths_health.final.var.1 <- logdeaths_health.final.summary$which
logdeaths_health.final.var <- colnames(logdeaths_health.final.var.1)[logdeaths_health.final.var.1[opt.size, ]][-1] #take the intercept put
logdeaths_health.final.var <- logdeaths_health.final.var[-c(1:(forced_size-1))]
logdeaths_health.final.var

# Final model
logdeaths_health.final.fit <- lm(logdeaths_per100K ~., 
                              logdeaths_health.sub[, c("logdeaths_per100K", "state_name", logdeaths_health.final.var)])
summary(logdeaths_health.final.fit)

plot(logdeaths_health.final.fit, 1)
plot(logdeaths_health.final.fit, 2)

```


[end]

